\textbf{universal corpus}

### 2014-04-24 docs/computEL-final/clean.tex
Minor changes after readthrough (typos, etc.)

\paragraph{UDHR.} Although there is a pre-compiled version of the UDHR data from the
Natural Language ToolKit (NLTK) corpora
distribution,\footnote{\url{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}}
the distribution is laden with encoding problems 
%during their
%conversion from pdf to plaintext format. 
Instead, we use the
plaintext files available from the Unicode
website,\footnote{\url{http://unicode.org/udhr/d}} which are free of
encoding issues. The first four lines of each file record metadata,
and the rest is the translation of the UDHR. This dataset is
extremely clean, and simply required segmentation into sentences.

\paragraph{Wikipedia.}
One major issue with using the Wikipedia dump is the problem of separating text from abundant source-specific markup. To convert compressed
Wikipedia dumps to textfiles, we used the WikiExtractor
\footnote{\url{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor}}
tool. After conversion into textfiles, we used several regular expressions to
delete residual Wikipedia markup and so-called ``magic
words"\footnote{\url{http://en.wikipedia.org/wiki/Help:Magic\_words}}.

\paragraph{Omniglot.}
The main issue with extracting this data is that the pages are designed to be human-readable, not machine-readable.  Cleaning this data required parsing the html source, and extracting the relevant content, which required different code for the two types of page we considered (useful phrases, and the Tower of Babel story).  Even after automatic extraction, some noise in the data remained, such as explanatory notes given in parentheses, which are written in English and not the target language.  Even though the total amount of data here is small compared to our ther sources, the amount of effort required to process it was not, because of these idiosyncracies.  We expect that researchers seeking to convert data from human-readable to machine-readable formats will encounter similar problems, but unfortunately there is unlikely to be a one-size-fits-all solution to this problem.

\paragraph{ODIN.}
The ODIN data is easily accessible in XML format from the online database,\footnote{\url{http://odin.linguistlist.org/download}} where data for each language is saved in a separate XML file and the IGTs are encoded in tags of the form \texttt{<igt><example>...</example></igt>}.  For example, the IGT in figure~\ref{fig:odin_fijian} is represented by the XML snippet in figure~\ref{fig:odin_fijian_xml}.

@@

\begin{figure}[t]
\quad 21 a.\quad o lesu mai \\
\indent \qquad\qquad 2sg return here \\
\indent \qquad\qquad `\emph{You return here.}' \\
\caption{Fijian IGT from ODIN.} \label{fig:odin_fijian}
\end{figure}

\begin{figure}[t]
\small
\begin{lstlisting}[language=XML]
<igt>
  <example>
    <line>21 a. o lesu mai</line>
    <line>2sg return here</line>
    <line>`You return here.'</line>
  </example>
</igt>
\end{lstlisting} 
\caption{Fijian IGT in ODIN's XML format.} \label{fig:odin_fijian_xml}
\end{figure}

@@

\begin{figure}[t]
\small
\begin{lstlisting}[language=XML]
<igt>
  <example>
    <line>(69) na-Na-tmi-kwalca-t 
    Yimas (Foley 1991)</line>
    <line>3sgA-1sgO-say-rise-PERF
    </line>
    <line>`She woke me up' 
    (by verbal action)</line>
  </example>
</igit>
\end{lstlisting} 
\smallskip
\caption{Yimas IGT in ODIN's XML format.}\label{fig:odin_yimas_xml}
\end{figure}

@@

Wikipedia uses its own set of language codes, most of which are in ISO~639-1 or ISO~639-3.  The older ISO~639-1 codes are easy to recognise, being two letters long instead of three, and can be straightforwardly converted.  However, a small number of Wikipedia codes are not ISO codes at all - we converted these to ISO~639-3, following documentation from the Wikimedia Foundation.\footnote{\url{http://meta.wikimedia.org/wiki/Special\_language\_codes}}

@@

It is important to make sure that the data we have compiled will be available to future researchers, regardless of how the surrounding infrastructure changes. \newcite{bird2003port} describes a set of best practices for maintaining portability of digital information, outlining seven dimensions along which this can vary. Following this advice, we have ensured that all our data is available as plain text files, with utf-8 encoding, labelled with the relevant ISO 639-3 code. Metadata is stored separately. We have written an API to allow access to the data according to the guidelines of \newcite{abney2010universal}, who remain agnostic as to the specific form of data storage. If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according to these guidelines.

>>>>

\paragraph{UDHR.} Although there is a pre-compiled version of the UDHR data from the
Natural Language ToolKit (NLTK) corpora
distribution,\footnote{\url{http://nltk.googlecode.com/svn/trunk/nltk_data/index.xml}}
the distribution is laden with encoding problems. 
%during their
%conversion from pdf to plaintext format. 
Instead, we use the
plaintext files available from the Unicode
website,\footnote{\url{http://unicode.org/udhr/d}} which are free of
encoding issues. The first four lines of each file record metadata,
and the rest is the translation of the UDHR. This dataset is
extremely clean, and simply required segmentation into sentences.

\paragraph{Wikipedia.}
One major issue with using the Wikipedia dump is the problem of separating text from abundant source-specific markup. To convert compressed
Wikipedia dumps to textfiles, we used the WikiExtractor\footnote{\url{http://medialab.di.unipi.it/wiki/Wikipedia_Extractor}}
tool. After conversion into textfiles, we used several regular expressions to
delete residual Wikipedia markup and so-called ``magic
words"\footnote{\url{http://en.wikipedia.org/wiki/Help:Magic_words}}.

\paragraph{Omniglot.}
The main issue with extracting the Omniglot data is that the pages are designed to be human-readable, not machine-readable.  Cleaning this data required parsing the HTML source, and extracting the relevant content, which required different code for the two types of page we considered ('\emph{Useful foreign phrases}' and '\emph{Tower of Babel}').  Even after automatic extraction, some noise in the data remained, such as explanatory notes given in parentheses, which are written in English and not the target language.  Even though the total amount of data here is small compared to our other sources, the amount of effort required to process it was not, because of these idiosyncracies.  We expect that researchers seeking to convert data from human-readable to machine-readable formats will encounter similar problems, but unfortunately there is unlikely to be a one-size-fits-all solution to this problem.

\paragraph{ODIN.}
The ODIN data is easily accessible in XML format from the online database\footnote{\url{http://odin.linguistlist.org/download}}. Data for each language is saved in a separate XML file and the IGTs are encoded in tags of the form \texttt{<igt><example>...</example></igt>}.  For example, the IGT in figure~\ref{fig:odin_fijian} is represented by the XML snippet in figure~\ref{fig:odin_fijian_xml}.

@@

\begin{figure}[t]
\quad 21 a.\quad o lesu mai \\
\indent \qquad\qquad 2sg return here \\
\indent \qquad\qquad `\emph{You return here.}' \\
\caption{Fijian IGT from ODIN} \label{fig:odin_fijian}
\end{figure}

\begin{figure}[t]
\small
\begin{lstlisting}[language=XML]
<igt>
  <example>
    <line>21 a. o lesu mai</line>
    <line>2sg return here</line>
    <line>`You return here.'</line>
  </example>
</igt>
\end{lstlisting} 
\caption{Fijian IGT in ODIN's XML format} \label{fig:odin_fijian_xml}
\end{figure}

@@

\begin{figure}[t]
\small
\begin{lstlisting}[language=XML]
<igt>
  <example>
    <line>(69) na-Na-tmi-kwalca-t 
    Yimas (Foley 1991)</line>
    <line>3sgA-1sgO-say-rise-PERF
    </line>
    <line>`She woke me up' 
    (by verbal action)</line>
  </example>
</igit>
\end{lstlisting} 
\smallskip
\caption{Yimas IGT in ODIN's XML format}\label{fig:odin_yimas_xml}
\end{figure}

@@

Wikipedia uses its own set of language codes, most of which are in ISO~639-1 or ISO~639-3.  The older ISO~639-1 codes are easy to recognise, being two letters long instead of three, and can be straightforwardly converted.  However, a small number of Wikipedia codes are not ISO codes at all - we converted these to ISO~639-3, following documentation from the Wikimedia Foundation.\footnote{\url{http://meta.wikimedia.org/wiki/Special_language_codes}}

@@

It is important to make sure that the data we have compiled will be available to future researchers, regardless of how the surrounding infrastructure changes. \newcite{bird2003port} describe a set of best practices for maintaining portability of digital information, outlining seven dimensions along which this can vary. Following this advice, we have ensured that all our data is available as plain text files, with utf-8 encoding, labelled with the relevant ISO 639-3 code. Metadata is stored separately. We have written an API to allow access to the data according to the guidelines of \newcite{abney2010universal}, who remain agnostic as to the specific form of data storage. If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according to these guidelines.

### 2014-04-24 docs/computEL-final/data.tex
Minor changes after readthrough (typos, etc.)

The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations in 1948, and represents the first global expression of human rights. It consists of 30 articles, amounting to about four pages of text. This is a useful document for CL, since it has been translated into a wide variety of languages, providing a highly parallel text.

@@

Wikipedia\footnote{\url{http://www.wikipedia.org}} is a collaboratively-edited encyclopedia, appealing to use for NLP because of its large size and easy availability. At the time of writing, it contained 30.8 million articles in 287 languages, which provides a sizeable amount of monolingual text in a fairly wide range of languages. Text dumps are made regularly, and can be downloaded from \url{www.dumps.wikimedia.org}.

@@

Secondly, it is unclear how to represent the gloss.  Since the gloss line is not written in a natural language, we cannot treat it as a simple translation.  However, it is not straightforward to incorporate it into the proposed structure for analysed texts, either.  
% We could simply always map the gloss to the \textsc{gloss} field, but this would run against the suggestion that this field should "contain symbols from controlled vocabularies".  Indeed, this information might be more suited to the \textsc{morph, pos,} or \textsc{rel} fields, but it would be difficult to automatically process ODIN to determine which morphemes should be assigned to which field.
One possible resolution is to move all elements of the gloss written in capital letters to the \textsc{morph} field (as functional elements are usually annotated in this way), and all remaining elements to the \textsc{gloss} field.  However, this loses information, since we no longer know which morpheme has which meaning.  To keep all information encoded in the IGT, we need to modify the \newcite{abney2011data}'s proposal.

@@

One benefit of this approach is that glossed texts are treated in exactly the same way as parallel texts.  There is a unique identifier for each morpheme, and glosses are stored under this identifier and the corresponding gloss code.  Furthermore, to motivate the important place of parallel texts in a universal corpus, Abney and Bird view translations into a high-resource reference language as a convenient surrogates of meaning. By the same reasoning, we can use glosses to provide a more detailed surrogate of meaning, only written in a metalanguage instead of a natural one.

>>>>

The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations in 1948, and represents the first global expression of human rights. It consists of 30 articles, amounting to about four pages of text. This is a useful document for NLP, since it has been translated into a wide variety of languages, providing a highly parallel text.

@@

Wikipedia\footnote{\url{http://www.wikipedia.org}} is a collaboratively-edited encyclopedia, appealing to use for NLP because of its large size and easy availability. At the time of writing, it contained 30.8 million articles in 286 languages, which provides a sizeable amount of monolingual text in a fairly wide range of languages. Text dumps are made regularly available, and can be downloaded from \url{www.dumps.wikimedia.org}.

@@

Secondly, it is unclear how to represent the gloss.  Since the gloss line is not written in a natural language, we cannot treat it as a simple translation.  However, it is not straightforward to incorporate it into the proposed structure for analysed texts, either.  
% We could simply always map the gloss to the \textsc{gloss} field, but this would run against the suggestion that this field should "contain symbols from controlled vocabularies".  Indeed, this information might be more suited to the \textsc{morph, pos,} or \textsc{rel} fields, but it would be difficult to automatically process ODIN to determine which morphemes should be assigned to which field.
One possible resolution is to move all elements of the gloss written in capital letters to the \textsc{morph} field (as functional elements are usually annotated in this way), and all remaining elements to the \textsc{gloss} field.  However, this loses information, since we no longer know which morpheme has which meaning.  To keep all information encoded in the IGT, we need to modify \newcite{abney2011data}'s proposal.

@@

One benefit of this approach is that glossed texts are treated in exactly the same way as parallel texts.  There is a unique identifier for each morpheme, and glosses are stored under this identifier and the corresponding gloss code.  Furthermore, to motivate the important place of parallel texts in a universal corpus, Abney and Bird view translations into a high-resource reference language as convenient surrogates of meaning. By the same reasoning, we can use glosses to provide a more detailed surrogate of meaning, only written in a metalanguage instead of a natural one.

### 2014-04-24 docs/computEL-final/relwork.tex
Minor changes after readthrough (typos, etc.)

Such resources are invaluable for many purposes
for instance, there can be plenty of material to work with for a
researcher who is looking to find data on a specific language, who is
willing to put in a little time browsing through different catalogues,
and who only needs human-readable data.  However, for large-scale
machine processing, they leave much to be desired.

@@

HC Corpora currently provides download of corpora in 68 different language varieties, which vary in size from 2M to 150M words. The corpora are thus of a respectable size, but only 1\% of the world's languages are represented.  A further difficulty is that languages are named, without the corresponding ISO codes. This problem may be small when all 68 varieties are well-known, but will become a serious concern if this corpus collection is to grow and include smaller languages.

>>>>

Such resources are invaluable for many purposes.
For instance, there can be plenty of material to work with for a
researcher who is looking to find data on a specific language, who is
willing to put in a little time browsing through different catalogues,
and who only needs human-readable data.  However, for large-scale
machine processing, they leave much to be desired.

@@

HC Corpora currently provides download of corpora in 68 different language varieties, which vary in size from 2M to 150M words. The corpora are thus of a respectable size, but only 1\% of the world's languages are represented.  A further difficulty is that languages are named, without the corresponding ISO language codes. This problem may be small when all 68 varieties are well-known, but will become a serious concern if this corpus collection is to grow and include smaller languages.
